---
title: "Introduction to cell2spatial Package"
author: "IPF Research Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cell2spatial Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `cell2spatial` package provides comprehensive tools for analyzing spatial transcriptomics data, with a particular focus on Idiopathic Pulmonary Fibrosis (IPF) research. This vignette demonstrates the key functions and workflows.

## Installation

```{r installation, eval=FALSE}
# Install from GitHub (when available)
devtools::install_github("your-username/cell2spatial")

# Load the package
library(cell2spatial)
```

```{r load-package}
library(cell2spatial)
```

## Core Functions Overview

The package includes several key functions:

- `t_test_compare_cell()`: Cell type-specific differential expression with optional FDR correction
- `wilcox_compare_region()`: Wilcoxon tests for region comparisons with rank-biserial correlation
- `t_test_compare_region()`: t-tests for region comparisons with Cohen's d effect sizes
- `calculate_signature_overlap()`: Gene set overlap enrichment analysis using hypergeometric test
- `calculate_multiple_overlaps()`: Batch overlap analysis across multiple region comparisons
- `calculate_signature_score()`: Calculate enrichment scores for gene signatures
- `calculate_multiple_signature_scores()`: Batch signature scoring across multiple signatures

### Multiple Testing Correction

All statistical comparison functions support flexible multiple testing correction via the `correction_method` parameter:

- **`"Nominal"`** (default): Returns unadjusted p-values
- **`"BH"`**: Applies Benjamini-Hochberg FDR correction

This gives you control over when and how to apply multiple testing correction.

## Example 1: Basic data setup

Prepare your spatial transcriptomics data for analysis:

```{r data-setup}
# Load example spatial transcriptomics data
# In practice, load your GeoMx or other spatial data

# Create example expression matrix
set.seed(123)
n_genes <- 200
n_roi <- 40

spatial_expression <- matrix(
  rnorm(n_genes * n_roi, mean = 5, sd = 2), 
  nrow = n_genes, 
  ncol = n_roi
)
rownames(spatial_expression) <- paste0("Gene_", 1:n_genes)
colnames(spatial_expression) <- paste0("ROI_", 1:n_roi)

# Create spatial metadata with region annotations
spatial_metadata <- data.frame(
  roi_id = paste0("ROI_", 1:n_roi),
  region_type = rep(c("foci", "alveoli", "normal"), c(15, 15, 10)),
  patient_id = rep(paste0("Patient_", 1:8), each = 5),
  diagnosis = rep("IPF", n_roi)
)

# View data structure
cat("Expression matrix dimensions:", dim(spatial_expression), "\n")
head(spatial_metadata)
```

## Example 2: Region specific gene analysis

Compare gene expression between spatial regions using non-parametric or parametric tests:

```{r region-comparison}
# Create simulated spatial region data
set.seed(567)
n_genes <- 100
n_samples <- 60

region_expression <- matrix(
  rnorm(n_genes * n_samples), 
  nrow = n_genes, 
  ncol = n_samples
)
rownames(region_expression) <- paste0("Gene_", 1:n_genes)
colnames(region_expression) <- paste0("S", 1:n_samples)

# Create sample metadata with regions
region_metadata <- data.frame(
  sample_id = paste0("S", 1:n_samples),
  region_type = rep(c("foci", "alveoli", "fibrosis"), each = 20)
)

# Wilcoxon test with nominal p-values
wilcox_results <- wilcox_compare_region(
  expression_data = region_expression,
  sample_metadata = region_metadata,
  target_region = "foci",
  correction_method = "Nominal"
)

head(wilcox_results[order(wilcox_results$p.value), ])

# t-test with FDR correction
ttest_results <- t_test_compare_region(
  expression_data = region_expression,
  sample_metadata = region_metadata,
  target_region = "foci",
  correction_method = "BH"
)

cat("\nTop genes from t-test (FDR corrected):\n")
head(ttest_results[order(ttest_results$p.value), ])
```

## Example 3: Query signature in single-cell data

After identifying region-specific genes from spatial data, query their expression in single-cell data:

```{r query-signature}
# Use significant genes from region analysis as signature
region_signature <- rownames(wilcox_results)[wilcox_results$p.value < 0.05]

# Create simulated single-cell expression data
set.seed(678)
n_genes_sc <- 500
n_cells <- 100

sc_expression <- matrix(
  rnorm(n_genes_sc * n_cells), 
  nrow = n_genes_sc, 
  ncol = n_cells
)
rownames(sc_expression) <- paste0("Gene_", 1:n_genes_sc)
colnames(sc_expression) <- paste0("Cell_", 1:n_cells)

# Create single-cell metadata
sc_metadata <- data.frame(
  cell_id = colnames(sc_expression),
  cell_type = rep(c("Fibroblasts", "Epithelial", "Immune"), c(40, 40, 20))
)

# Calculate signature scores for each cell type
signature_scores <- calculate_signature_score(
  expression_data = sc_expression,
  sample_metadata = sc_metadata,
  signature_genes = region_signature
)

# View results
head(signature_scores)

# Compare scores across cell types
cat("\nMean signature scores by cell type:\n")
aggregate(score ~ cell_type, data = signature_scores, FUN = mean)
```

## Example 4: Cell type specific gene analysis

Identify cell type-specific marker genes from single-cell data:

```{r cell-type-example}
# Create simulated expression matrix
set.seed(456)
n_genes <- 100
n_samples <- 50

expression_matrix <- matrix(
  rnorm(n_genes * n_samples), 
  nrow = n_genes, 
  ncol = n_samples
)
rownames(expression_matrix) <- paste0("Gene_", 1:n_genes)
colnames(expression_matrix) <- paste0("Sample_", 1:n_samples)

# Create metadata
metadata <- data.frame(
  celltype = rep(c("Fibroblasts", "Epithelial", "Immune"), c(20, 20, 10)),
  Diagnosis = rep("IPF", n_samples),
  row.names = colnames(expression_matrix)
)

# Perform cell type analysis with nominal p-values (default)
cell_results <- t_test_compare_cell(
  normalized_count = expression_matrix,
  meta_data = metadata,
  target_cell_type = "Fibroblasts",
  diagnosis = "IPF",
  correction_method = "Nominal"
)

# View top results
head(cell_results[order(cell_results$stat.sig), ])

# Alternative: Apply Benjamini-Hochberg FDR correction
cell_results_fdr <- t_test_compare_cell(
  normalized_count = expression_matrix,
  meta_data = metadata,
  target_cell_type = "Fibroblasts",
  diagnosis = "IPF",
  correction_method = "BH"
)

# Compare number of significant genes
cat("Significant genes (nominal p < 0.05):", 
    sum(cell_results$stat.sig < 0.05), "\n")
cat("Significant genes (FDR < 0.05):", 
    sum(cell_results_fdr$stat.sig < 0.05), "\n")
```

## Example 5: Overlap enrichment analysis

Test enrichment between region-specific and cell type-specific gene signatures:

```{r overlap-example}
# Create simulated gene lists for region-specific genes
set.seed(789)
all_genes <- paste0("Gene_", 1:1000)
foci_genes <- sample(all_genes, 200)
alveoli_genes <- sample(all_genes, 150)

# Perform overlap enrichment between two region signatures
overlap_result <- calculate_signature_overlap(
  region_signature_genes = foci_genes,
  query_signature_genes = alveoli_genes,
  universe_size = length(all_genes)
)

# Print key results
cat("Overlap count:", overlap_result$overlap, "\n")
cat("P-value:", overlap_result$p.value, "\n")
cat("Enrichment type:", overlap_result$enrichment_type, "\n")
```

## Example 6: Multiple overlap analysis

For analyzing multiple region signatures:

```{r multiple-overlap-example}
# Create multiple region gene signatures
region_signatures <- list(
  foci = sample(all_genes, 150),
  alveoli = sample(all_genes, 120),
  fibrosis = sample(all_genes, 100)
)

reference_signature <- sample(all_genes, 200)

# Perform multiple overlap analysis across regions
multiple_results <- calculate_multiple_overlaps(
  region_signature_list = region_signatures,
  query_signature_genes = reference_signature,
  universe_size = length(all_genes)
)

print(multiple_results)
```


## Function Parameters

### wilcox.comp() Parameters

- `data`: Data frame with expression and grouping data
- `group_var`: Column name for grouping variable
- `target_group`: Specific group to compare
- `expression_var`: Column name for expression values
- `alternative`: Test alternative ("two.sided", "greater", "less")

### t_test_compare_cell() Parameters

- `normalized_count`: Expression matrix (genes × samples)
- `meta_data`: Sample metadata data frame
- `target_cell_type`: Cell type of interest
- `diagnosis`: Diagnosis condition to analyze
- `min_samples`: Minimum samples required (default: 3)
- `correction_method`: Multiple testing correction - "Nominal" (default) or "BH"

### wilcox_compare_region() Parameters

- `expression_data`: Expression matrix (genes × samples)
- `sample_metadata`: Sample metadata data frame
- `target_region`: Target spatial region for comparison
- `sample_col`: Sample ID column (optional)
- `region_col`: Region type column (optional)
- `correction_method`: Multiple testing correction - "Nominal" (default) or "BH"

### t_test_compare_region() Parameters

- `expression_data`: Expression matrix (genes × samples)
- `sample_metadata`: Sample metadata data frame
- `target_region`: Target spatial region for comparison
- `sample_col`: Sample ID column (optional)
- `region_col`: Region type column (optional)
- `var_equal`: Assume equal variances (default: FALSE for Welch's t-test)
- `correction_method`: Multiple testing correction - "Nominal" (default) or "BH"

### calculate_signature_overlap() Parameters

- `region_signature_genes`: Gene symbols from region-specific signature
- `query_signature_genes`: Gene symbols from query signature  
- `universe_size`: Total number of genes in background universe

### calculate_multiple_overlaps() Parameters

- `region_signature_list`: Named list of region gene signature vectors
- `query_signature_genes`: Query signature gene symbols
- `universe_size`: Total number of genes in background universe

## Tips and Best Practices

1. **Data Quality**: Ensure your expression data is properly normalized
2. **Sample Size**: Use adequate sample sizes for reliable statistics
3. **Multiple Testing Correction**: 
   - Use `correction_method = "Nominal"` if you want to apply custom correction later
   - Use `correction_method = "BH"` for automatic Benjamini-Hochberg FDR correction
   - For exploratory analysis, consider using nominal p-values first
4. **Effect Sizes**: Always report effect sizes (Cohen's d, rank-biserial) alongside p-values
5. **Test Selection**:
   - Use Wilcoxon tests for non-normal distributions or small sample sizes
   - Use t-tests when assumptions of normality are met
   - Welch's t-test (var_equal = FALSE) is more robust to unequal variances
6. **Background Sets**: Use appropriate background gene sets for enrichment analysis
7. **Validation**: Validate results with independent datasets when possible

## Getting Help

For more information about specific functions:

```{r help-example, eval=FALSE}
?t_test_compare_cell
?wilcox_compare_region
?t_test_compare_region
?calculate_signature_overlap
?calculate_signature_score
```

## Summary of All Package Functions

### Statistical Comparison Functions

| Function | Purpose | correction_method | Effect Size/Statistic |
|----------|---------|-------------------|----------------------|
| `t_test_compare_cell()` | Cell type differential expression (Welch's t-test) | "Nominal" or "BH" | Cohen's d |
| `wilcox_compare_region()` | Non-parametric region comparison (Wilcoxon) | "Nominal" or "BH" | Rank-biserial correlation |
| `t_test_compare_region()` | Parametric region comparison (t-test) | "Nominal" or "BH" | Cohen's d |

### Overlap Enrichment Functions

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `calculate_signature_overlap()` | Single overlap enrichment analysis | 2 gene lists + universe size | Hypergeometric p-value, overlap count |
| `calculate_multiple_overlaps()` | Batch overlap analysis across regions | Multiple gene lists + query | Data frame with all comparisons |

### Gene Signature Scoring Functions

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `calculate_signature_score()` | Score single gene signature | Expression matrix + gene list | Enrichment scores per sample |
| `calculate_multiple_signature_scores()` | Score multiple signatures at once | Expression matrix + signature list | Score matrix (samples × signatures) |
| `query_signature_comprehensive()` | Comprehensive signature query | Expression matrix + signature + metadata | Detailed signature analysis |

### Data Normalization Functions

| Function | Purpose | Method | Output |
|----------|---------|--------|--------|
| `normalization_by_gene()` | Gene-wise normalization | Z-score transformation | Normalized expression matrix |
| `normalize_expression()` | Matrix normalization | Z-score (default) or other methods | Normalized matrix |

### Visualization Functions

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `plot_signature_scores()` | Main signature visualization | Signature score matrix | ggplot2 bubble plot |
| `plot_positive_signature()` | Visualize positive signatures | Signature score matrix | ggplot2 plot |
| `plot_negative_signature()` | Visualize negative signatures | Signature score matrix | ggplot2 plot |

### Helper/Internal Functions

- `calculate_rank_biserial_correlation()` - Effect size for Wilcoxon test
- `calculate_cohens_d()` - Effect size for t-test

## Session Info

```{r session-info}
sessionInfo()
```